---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

 

```{r load-data}
load("brfss2013.RData")
```



* * *

## Part 1: Data
### Brief information about BFRSS
The Behavioral Risk Factor Surveillance System (BRFSS) is a collaborative project between all of the states in the United States (US) and participating US territories and the Centers for Disease Control and Prevention (CDC). The BRFSS is administered and supported by CDC's Population Health Surveillance Branch, under the Division of Population Health at the National Center for Chronic Disease Prevention and Health Promotion. BRFSS is an ongoing surveillance system designed to measure behavioral risk factors for the non-institutionalized adult population (18 years of age and older) residing in the US.

### The mission of BFRSS
The BRFSS objective is to collect uniform, state-specific data on preventive health practices and risk behaviors that are linked to chronic diseases, injuries, and preventable infectious diseases that affect the adult population. Factors assessed by the BRFSS in 2013 include tobacco use, HIV/AIDS knowledge and prevention, exercise, immunization, health status, healthy days - health-related quality of life, health care access, inadequate sleep, hypertension awareness, cholesterol awareness, chronic health conditions, alcohol consumption, fruits and vegetables consumption, arthritis burden, and seatbelt use.

### Data collection
Since 2011, BRFSS conducts both landline telephone- and cellular telephone-based surveys. In conducting the BRFSS landline telephone survey, interviewers collect data from a randomly selected adult in a household. In conducting the cellular telephone version of the BRFSS questionnaire, interviewers collect data from an adult who participates by using a cellular telephone and resides in a private residence or college housing. Therefore, BRFSS uses 2 separate samples for conducting the research: 1) The landline sample and 2) The cellular telephone sample

#### The landline sample
For the landline sample, Disproportionate stratified sampling (DSS) has been used, which means the sample size of each stratum does not have to be proportional to the population size of the stratum.Telephone numbers are drawn from two strata that are based on the presumed density of known telephone household numbers. In this design, telephone numbers are classified into strata that are either high density or medium density to yield residential telephone numbers. Telephone numbers in the high
density stratum are sampled at the highest rate.

#### The cellular telephone sample
The cellular telephone sample is randomly generated from a sampling frame of confirmed cellular area code and prefix combinations. Cellular telephone respondents are randomly selected with each having equal probability of selection. States complete approximately 20% of their completed interviews with respondents on cell phones.

### Implications of data collection method on scope of inference
There are some drawbacks in sampling method and the study method itself which can prevent drawing strong conclusions from the study in general.

First of all, using telephones for conducting a study introduces some bias. In this way, study limits the sampling to those who have a landline telephone and reside in residential or college housing or have cellular telephone. Additionally, there is possibility that even the ones who have telephones may not answer the call made by study conductors which introduces nonresponse bias.

Secondly, the samples sizes from each state are not proportional to state population (not normalized), therefore small and large states are not represented according to their population size.

The last but not least is the issue with the interview format. Respondents themselves may not know/remember accurate answer to some questions asked such as, During the past 30 days, for about how may days have you felt sad, blue, or depressed?  or they could intentionally give wrong answers due to the fact that some of the questions are personal or sensitive, such as Days In Past 30 Had Alcoholic Beverage.

#### Generalizability
Sampling aimed at non-institutionalized adult population (18 years of age and older) residing in the US. Overall, an estimated 97.5% of US households had telephone service in 2012 and telephone coverage varies across states with a range of 95.3% in New Mexico to 98.6% in Connecticut. As noted in Data collection section, stratified sampling methodology is used for landline telephone sample, and random sampling for cellular telephone sample.  Due to large sample size, extensive areal coverage and random sampling method, the study can be generalized to all non-institutionalized adult population residing in the US.

#### Causality
As the study is observational in nature and there is not any random assignment to study groups, the survey cannot be used to make causal relationships. limiting an inference scope to associative relationships.


* * *

## Part 2: Research questions

**Research quesion 1:**
Is there is any correlation between number of days person feel depressed in past 30 days and number of times he/she did exercise in past 30 days? Is this relationship different for male and females?

This research question would help to determine whether doing exercises have an association with stress level and how different this association is between 2 sexes ?



**Research quesion 2:**
How does number of hours spent on job associates with $75 000+ income in each state?

This question aims to find whether there is a correlation between working long hours and income level and how is it different in each state?


**Research quesion 3:**
Does amount of time spent on sleep associates with the feeling of fully energized (past 30 days)?

Here aim is to determine whether amount of sleep correlates with the feeling of energized.


* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Research quesion 1:**
Is there is any correlation between number of days person feel depressed in past 30 days and number of times he/she did exercise in past 30 days? Is this relationship different for male and females?

Let's first rename qlmentl2 variable ,which represents the number of days person feel depressed in past 30 days, to days_felt_stressed variable and exerany2, which shows whether the person exercised or not during last 30 days, to did_exercise_or_not variable. As we renamed both variables, we can now explore them

```{r}
brfss2013 = rename(brfss2013, did_exercise_or_not = exerany2, days_felt_stressed = qlmentl2)
brfss2013 %>% group_by(days_felt_stressed) %>% summarize(count=n())

```
We see that for most of the respondents the value is missing (NA), which dramatically reduces the number of observations. Excluding the observations with missing value, majority of people reported 0 days as number of days they felt depressed.

```{r}
brfss2013 %>% group_by(did_exercise_or_not) %>% summarize(count=n())

```
Here, fortunately we see that the number of missing values are quite lower and most respondents reported that they did exercises in last 30 days.

Now, we can extract days_felt_stressed, did_exercise_or_not and sex variables from original dataset and make new dataframe for further data analysis and visualization and remove all missing values. 

```{r}
research1 <- subset(brfss2013, select=c("days_felt_stressed", "did_exercise_or_not","sex" ))
research1 <- na.omit(research1)
```

We can subset data do 2 separate bar charts to compare results between males and females:
```{r}
research1_male <- subset(research1, sex=="Male")
research1_female <- subset(research1, sex=="Female")
ggplot(data=research1_male, aes(days_felt_stressed, fill=did_exercise_or_not)) +
geom_bar(stat="bin")+labs(title = "Bar chart for males data")
ggplot(data=research1_female, aes(days_felt_stressed, fill=did_exercise_or_not)) +
geom_bar(stat="bin")+labs(title = "Bar chart for females data")
```


According to 2 bar charts there is not any strong association between doing exercises and number of days felt stressed for both males and females. However, as data on doing exercises is categorical and it is not possible to determine whether "Yes" means doing exercise once for last 30 days or at higher extreme every day. Therefore, based on our data it is not readily possible to determinate correlation between physical activeness and number of days felt stressed.

**Research quesion 2:**
How does number of hours spent on job associates with $75 000+ income in each state?

Again some data transformation and visualization

```{r}
brfss2013 = rename(brfss2013, state = X_state, number_of_hours_work =scntwrk1,income=income2)
research2 <- subset(brfss2013, select=c("state", "number_of_hours_work","income" ))
research2 <- na.omit(research2)
research2 <- subset(research2, income=="$75,000 or more")
average_by_state <-aggregate(research2[,2], list(research2$state), mean)
average_by_state = rename(average_by_state, state = Group.1, average_hours_worked=x)
average_by_state$state <- factor(average_by_state$state, levels = average_by_state$state[order(average_by_state$average_hours_worked)])
ggplot(data=average_by_state, aes(x=state, y=average_hours_worked), las=2) +
geom_bar(stat="identity")+theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))+labs(title = "Average number of hours spent on work in each state for people with $75 000 + income")
```

Unfortunately, after removing missing values from data and filtering income to be $75 000+, we missed most of the data.
To answer our research question we first find how many hours on average people with $75 000+ spent to job in each state. Visualized sorted bar chart shows that, although there is not much difference in average number of hours spent to job for people with $75 000+ income across states, maintaining $75 000+ income in Louisiana correlates with longer number of hours in comparison to other states

**Research quesion 3:**
Does amount of time spent on sleep associates with the feeling of fully energized (past 30 days)?

```{r}
brfss2013 = rename(brfss2013, sleep_time=sleptim1, days_full_energy=qlhlth2)
research3 <- subset(brfss2013, select=c("sleep_time", "days_full_energy"))
research3 <- na.omit(research3)
ggplot(research3, aes(x=sleep_time, y=days_full_energy)) + geom_point()+geom_smooth(method=lm)+labs(title = "        Days people felt energized vs sleep time")

```

Although due to discrete nature of numerical variables, visualized scatterplot is far from ideal, in general after inserting regression line we see that overall there is positive correlation between sleep time and days people felt energized.

